Clinical Deployment Environments: The Five Pillars of Translational Machine Learning for Health

### 2022-06-29 rebuttals/response-reviewer-2.md

I might suggest omitting or significantly abbreviating this section, which could open space to insert valuable information from the supplement. I also think a paragraph commenting on the relationship of the CDE to the organization's "problem identification and specification" apparatus would be a very helpful "bookend" to the implementation and continuous evaluation portions of the essay.

- [ ] ask Tim to think how he wants to play this?
https://github.com/docsteveharris/2022-adversarial-penguin/issues/5

2. Some areas of the prose could benefit from increased specificity and/or clarity. One of the challenges in the healthcare data science space is striking the right balance of technical language to meet a readership with different levels of familiarity to clinical medicine, coding/development, and operations. Some areas of the prose - e.g., lines 63-67 - skew a bit heavy into the data science jargon and might benefit from additional explanations or a bit less jargon. It may be worth the authors having a colleague with less technical familiarity provide some informal copy-editing to help with readability for an audience of varying levels of expertise.

Thank you. This was helpful. I have asked two colleagues to separately proof the manuscript. They have made a series of small stylistic changes that I have incorporated into the manuscript.

e.g. lines 63-67

Before:

> Given the sensitive nature of health data, these initiatives depend on expanding investment into Trusted Research Environments (TRE). [@2021c] Data flows from source (primary, secondary, social care and elsewhere) to a single secure landing zone where research teams write the code to link, clean and analyse the data. The insights return to the bedside through clinical guidelines and policy. This offline '_data-to-code_' approach is also the dominant design pattern in ML4H projects but is fundamentally flawed.

After

> Given the sensitive nature of health data, these programmes depend on Data Safe Havens (also known as Trusted Research Environments [TRE]). [@2021c] TREs are an example of 'data-to-modeler' (DTM) designs where source data (from primary, secondary, and social care) **moves to** a separate, secure landing zone. Here research teams link, clean and analyse the data. Their insights eventually return to the bedside through clinical guidelines and policy. In ML4H, DTM is also the dominant design pattern but is fundamentally flawed.

A detailed diff of these changes is available [here](https://github.com/docsteveharris/2022-adversarial-penguin/pull/7)

### 2022-06-29 rebuttals/response-reviewer-1.md

# Independent Review Report, Reviewer 1

The article proposes a design pattern, Clinical Deployment Environments (CDE) with 05 principles (pillars). A design pattern is a general repeatable and implementable solution which is well demonstrated through the datasheet attached. My overall impression apart from a few specific comments is, that for each of the five principles (pillars), the problem description that they address may be clearly stated in the introduction section. The implementation example in the datasheet describes the first 02 pillars very clearly, however, an implementation example for the rest of the 03 pillars is missing. They should be included to prove the implementation. I would think that the Implementation science is a result of the first two pillars (i.e. RW-Dev and ML-Ops) and not essentially one of the pillars here. The explanation of 'Implementation Science' in the article also does not provide clear requirements for this pillar.

Other specific comments -
- The language style preferred for Frontiers in Digital Health Articles is US English while the article uses the GB English style. A revision is needed to match the editorial requirements

Thank you. We have converted converted the article to Word, and applies the US American spell check and grammar, and then converted back to Latex.

- A review of the state-of-the-art in the area may be included to highlight the gaps

Thank you. We have added several new references.

General

1. Feng J, Phillips RV, Malenica I, Bishara A, Hubbard AE, Celi LA, et al. Clinical artificial intelligence quality improvement: towards continual monitoring and updating of AI algorithms in healthcare. npj Digit Med. 2022 Dec;5(1):66. 
2. Eaneff S, Obermeyer Z, Butte AJ. The Case for Algorithmic Stewardship for Artificial Intelligence and Machine Learning Technologies. JAMA. 2020 Oct 13;324(14):1397–8. 
3. Lee CS, Lee AY. Clinical applications of continual learning machine learning. The Lancet Digital Health. 2020 Jun 1;2(6):e279–81.

Specific
<!-- TODO -->

- A thorough check for sentences, the flow of the information, and grammar will help increase the readability of the article. A few of the comments and errors are highlighted in the attached file.

Thank you. We have corrected as follows
1. Highlight [page 3]:* and aims to answer for ML4H the same challenge that translational medicine brought to bear for drug discovery *and Note [page 3]:* The sentence looks grammatically incomplete

**Re-written**:

> Just as bio-medicine identified translational roadblocks in the bench-to-bedside journey, ML4H also requires a translational infrastructure supporting real-world development not just real-world data.

2. *Highlight [page 7]:* (T2)*and Note [page 7]:* There is no mention or explanation of T2/T1 in the whole article

**Note**

This was defined in paragraph 1 of Section 4.

> This corresponds to the second (T2) arm of translational medicine: implementation science.[@woolf2008b]

3. *Highlight [page 8]:* does not have to relevant in the next *and Note [page 8]:* does not have to be relevant in the next

 **Thank you. Amended.**

 4.  *Highlight [page 8]:* Excellent and senior patient*and Note [page 8]:* Do you mean excellent patient and public involvement?

 **Yes. Thank you. Amended.**

 5. *Highlight [page 8]:* At our own institution, *and Note [page 8]:* Would you like to move this part as part of Data Sheet (as implementation example)?

 We would be keen to keep this brief paragraph in the body of the main text. Our original draft had the full explanation. This short note was meant to provide enough detail to trigger a reader's interest without distracting from the broader emphasis of the article.

 6.  *Highlight [page 9]:* such binding*and Note [page 9]:* such as binding...

 **Thank you. Amended.**

7. Highlight [page 10]:* where software is (not as) a medical device *and Note [page 10]:* Do you mean Software as Medical Device?

No. We were attempting to point out that where software is within a medical device (e.g. a cardiac pacemaker) then we already have a good regulatory framework. But because ML4H will be deployed in less controlled and varying environment (the EHR) then we need a new approach.

8. *Strikeout [page 10]:* create

 **Thank you. Amended.**

- The diagram in Figure 1. does not match with the attached Image 1

I'm not sure if I understand. We have uploaded 2 copies of each diagram, and in our version I think they are numbered correctly (I have double checked). Could you please have another look. Thanks.

- Suggest moving the figures as part of the datasheet where EMAP and FlowEHR are explained instead of keeping them in the main article
- An overall depiction of CDE and its pillars may be useful

We're happy to take editorial guidance on this. Our inclination is to include the detailed diagrams in the main text and to leave the textual and more technical descriptions of Pillars 1 and 2 in the ESM. We think this keeps the main article readable as an overall 'sketch' but allows the interested reader to obtain further detail. To this end (as per Reviewer 3), we have added extended captions rather than extend the body text.

- Datasheet, Page 2, Section 1.4.2, "... aggregation. (201, 2018) Flat (bulk) FHIR is not mature, nor designed for live interactions.(202, 2021)..." : Please update the references. They are not numbered in the bibliography and full information is missing.

Thank you. These were internet (web) references, and we have corrected the missing details.

### 2022-06-29 rebuttals/response-reviewer-2.md

3. Much of the supplemental content would be better served in the main text.

- [ ] ditto; decide how to play this

## Major Comments
1. The analogy to drug discovery is a poor fit for this essay. One key premise underlying drug discovery activities is that potentially-actionable targets have already been identified through prior research. For example, the decision to seek JAK2-interacting molecules emerged from the understanding that JAK2 regulates important cytokines in a number of disease states. The algorithmic equivalent of this kind of prior knowledge is not apparent to me. Without a clear understanding of (a) a well-articulated problem to be solved and (b) a reasonable mechanism/logic model in which an ML tool - provided it performs well - might affect stakeholder behavior, translational failure seems essentially guaranteed. Said differently, the bake-off between different ML approaches can really only be held once the model ("ligand")'s "receptor" has been mapped/described, which I would argue happens infrequently at present. As I read the authors' proposed pillars, I found myself reflecting on the need for this kind of formative work, informed by stakeholders, to ensure that the CDE can be used for properly-specified development, operations, and evaluation.

I might suggest omitting or significantly abbreviating this section, which could open space to insert valuable information from the supplement. I also think a paragraph commenting on the relationship of the CDE to the organization's "problem identification and specification" apparatus would be a very helpful "bookend" to the implementation and continuous evaluation portions of the essay.  
*We thank the reviewer for this comment but feel that changing the manuscript as they suggest would not be the right thing to do. The point of the drug development pipeline analogy is to articulate the complexity of the intellectual work required for innovation. No-one expects a drug to be developed from scratch in a 3-5 years. However, in many cases people are disappointed when healthcare AI algorithms do not yield practice-changing results in the timeframe of a single project. The benefit of the Drug Trial Phases is that people are easily able to understand the risk level, the size of trial and the appropriate endpoints. This same thinking needs to be related to algorithms. Similar points have been made by Pepe (2005) and Ferrante di Ruffano with respect to biomarker discovery.*
_To specifically address the JAK-2 example, this work would happen in the pre-clinical phase that we describe as algorithm discovery. The AI equivalent of JAK-2 (a targetable molecule) is a feature (a representation of data that has some utility in classification or prediction that also has a causal relationship with the phenomenon under study) _

@@

> Given the sensitive nature of health data, these programmes depend on Data Safe Havens (also known as Trusted Research Environments [TRE]). [@2021c] TREs are an example of 'data-to-modeler' (DTM) designs where source data (from primary, secondary, and social care) **moves to** a separate, secure landing zone. Here research teams link, clean and analyse the data. Their insights eventually return to the bedside through clinical guidelines and policy. In ML4H, DTM is also the dominant design pattern but is fundamentally flawed.

A detailed diff of these changes is available [here](https://github.com/docsteveharris/2022-adversarial-penguin/pull/7)

3. The supplement - especially items 1 and 2 - are very nice and highly practical. Some elements of these items could serve to address issue 2 above. Further, the ML-Ops design principles are particularly helpful explicating how code-to-data can address the limitations of data-to-code. I think it would be great if these could be a bit more prominent, ideally in the main manuscript.

Minor Comments
1. The manuscript has a number of instances with unspecified pronouns, e.g., sentences starting with "This" or "That," which do not specify what "This" or "That" actually are.

Thank you. We hope we have corrected this with the proof reading and recommendations (esp. reviewer 1).

2. Supplement page 2, typo in the ML-0PS header (zero instead of O)

### 2022-06-29 rebuttals/response-reviewer-3.md

Thank you. These references were enormously helpful, and we have included them in the text for Section 2.

> Others have made similar proposals and labelled this as "algorithmic stewardship" or "AI-QI".[@eaneff2020;@feng2022]

3. Recent work have questioned the value of model explainability, particularly because many explainability methods exhibit flaws. See Ghassemi, Oakden-Rayner, and Beam 2021 and [https://arxiv.org/abs/2202.01602](https://arxiv.org/abs/2202.01602), as well as references within. I think the recommendation for model explainability should be qualified with this fact that the current explainability methods leave much to be desired.

Thank you. We have revised the paragraph and added the recommended references as follows:

> We argue that model explainability methods [@gunning2019],[@mueller2019],[@vilone2020],[@Linardatos2020] need to be prioritised to help systematise and coordinate the processes of model troubleshooting by developers, risk-management by AI-backed service provider and system-checks by auditors.  Most AI models that operate as ‘black-box models’ are unsuitable for mission-critical domains, such as in healthcare, because they pose risk scenarios where problems that occur can remain masked and therefore undetectable and unfixable. As highlighted in [@Ghassemi 2021],[@Satyapriya2022] explainability methods cannot yet be relied on to provide a determinate answer as to whether an AI-recommendation is correct. However, explainability methods that highlight decision-relevant parts of AI representations and for measuring and benchmarking interpretability [@Doshi-Velez2017],[@Hoffman2018] are particularly promising for risk management as they can be used to structure a systematic interrogation of the trade-off between interpretability, model accuracy and the risk of model misbehaviour.

@@

4. There is a quickly growing literature on fail-safes in clinical ML and the authors should point readers to it. This idea has been referred to as selective prediction (Chow, 1970; Tortorella,2000; Bartlett and Wegkamp, 2008; El-Yaniv and Wiener, 2010; Feng, Sondhi, Perry, and Simon 2021). The idea of deploying ensembles with fail-safes is promising as well, and has also been previously described in Feng 2021.

Thank you. We have revised the paragraph and added the recommended references as follows:

> Fail safes, otherwise referred to as selective prediction [@Chow1970],[@ Bartlett 2008],[@Tortorella2000],[@ElYaniv2010],[@Feng2021] should be designed into support systems to pre-empt and mitigate model misbehaviour. The European Commission High-Level Expert Group on AI presented the Ethics Guidelines for Trustworthy Artificial Intelligence in April 2019 with recommendations for AI-support systems that continue to maintain human-agency via a human-in-the-loop oversight. Prediction models that map patient data to medically meaningful classes are forced to predict within the predetermined set of classes without the option to flag users when the model is unsure of an answer. To address this problem, there is good evidence that methods such as Bayesian deep learning and various uncertainty estimates [@abdar2021] can provide promising ways to detect and refer data samples with high probability of misprediction for human expert review [@Leibig2017; @Filos2019; @Ghoshal2020]. This may even permit less interpretable models to operate when implemented in conjunction with an effective fail-safe system.

@@

5. While I agree that dynamic model calibration is an important tool to maintain the reliability of ML systems, the first motivating example provided in Section 3.3 seem appropriate. The example states that an ML model can influence shifts in its input data, which can lead to performance deterioration. However, the authors also cite Liley, which shows an example where there are shifts induced by the model but the model itself should not be updated. The second motivating example (models trained on a bad dataset and therefore should be updated) is a good reason for model updating, though dynamic model calibration may not be sufficient in this case; instead larger model revisions may be required. Perhaps a better motivating example for dynamic model calibration is provided in Davis, Lasko, Chen, Siew, and Matheny 2017: They show how a ML model for acute kidney injury decays in model calibration.

Thank you. We have revised the paragraph and added the recommended references as follows:

> As discussed, models that influence the evolution of its own future input data are at risk of performance deterioration over time due to input data shifts [@Davis2017]. In such cases, continual learning via calibration drift detection and model recalibration [@Davis2020],[@Feng2022] provides a promising solution but remains a challenging paradigm in AI. Recalibration with non-stationary incremental data can lead to catastrophic forgetting when the new data negatively interferes with what the model has already learned [@Parisi2019], or a convergence where the model just predicts its own effect and thus should not be updated [@liley2021]. On the other hand, models can propose poor decisions because of the inherent biases found within the original dataset. In this case, dynamic model recalibration is unlikely to be sufficient and larger model revisions may be required. Here Pillar 1 (RW-dev) with suitable audit and monitoring via Pillar 2 (ML-Ops) will be required to overcome what would otherwise be a learning process encumbered by regulatory barriers.[@LEE2020]

@@

6. The authors do not provide references for how to perform dynamic model calibration. The following papers have proposed methods for doing so, and should be included in the manuscript: Davis, Greevy, Lasko, Walsh, Matheny 2020 and Feng, Gossmann, Sahiner, and Pirracchio 2022.

Thank you. We have included these additional references in the revised paragraph (quoted under (5) above).

### 2022-07-01 rebuttals/response-reviewer-1.md

# Independent Review Report, Reviewer 1

The article proposes a design pattern, Clinical Deployment Environments (CDE) with 05 principles (pillars). A design pattern is a general repeatable and implementable solution which is well demonstrated through the datasheet attached. My overall impression apart from a few specific comments is, that for each of the five principles (pillars), the problem description that they address may be clearly stated in the introduction section. The implementation example in the datasheet describes the first 02 pillars very clearly, however, an implementation example for the rest of the 03 pillars is missing. They should be included to prove the implementation. I would think that the Implementation science is a result of the first two pillars (i.e. RW-Dev and ML-Ops) and not essentially one of the pillars here. The explanation of 'Implementation Science' in the article also does not provide clear requirements for this pillar.

**Response:** Thank you. We hadn't previously considered the analogy to formal software design patterns ("a design or implementation structure that achieves a particular purpose") but this exactly captures our purpose. The manuscript is a conceptual over-view, and the ESM contains the detail where we have made progress ourselves. Pillars 3 and 4 are the requirements where we have made the least progress but removing them from the manuscript because of this would send the wrong message. We include them to acknowlege their importance and provide a series of useful references.

With regard to Implementation Science, we see this a distinct requirement that depends on Pillars 1 and 2 but is not captured by them. Indeed, it requires different personnel and different methodology. In most instances we imagine they will deliver their impact through some of form of passive (choice architecture) or active (interruptive alerts) decision support.

We have revised Section 4 to this end.

> Unlike medications, algorithms can only impact health by influencing the behaviour of clinicians and patients. This corresponds to the second (T2) arm of translational medicine: implementation science.[@woolf2008b] A well designed, safe, and responsible AI algorithm may still be ineffective if it does not reach a modifiable target on the clinical pathway.[@the2021] *In most instances, this will be via a modification of the choice architecture (passive)[@halpern2018; @main2010] or via interruptive alerts (active) embedded in the EHR.[@main2010]* Effective implementation requires a multi-disciplinary approach including human-computer interaction, behavioural science, and qualitative analysis.[@sendak2020]

Other specific comments -

- The language style preferred for Frontiers in Digital Health Articles is US English while the article uses the GB English style. A revision is needed to match the editorial requirements

**Response:** Thank you. We have converted converted the article to Word, and applies the US American spell check and grammar, and then converted back to Latex.

@@

**Response:** Thank you. We have added several new references.

General

1. Feng J, Phillips RV, Malenica I, Bishara A, Hubbard AE, Celi LA, et al. Clinical artificial intelligence quality improvement: towards continual monitoring and updating of AI algorithms in healthcare. npj Digit Med. 2022 Dec;5(1):66. 
2. Eaneff S, Obermeyer Z, Butte AJ. The Case for Algorithmic Stewardship for Artificial Intelligence and Machine Learning Technologies. JAMA. 2020 Oct 13;324(14):1397–8. 
3. Lee CS, Lee AY. Clinical applications of continual learning machine learning. The Lancet Digital Health. 2020 Jun 1;2(6):e279–81.

These are alongside additional references in Section 3 (AI safety), and Section 4 (Implementation).

@@

- The diagram in Figure 1. does not match with the attached Image 1

I'm not sure if I understand. We have uploaded 2 copies of each diagram, and in our version I think they are numbered correctly (I have double checked). Could you please have another look. Thanks.

- Suggest moving the figures as part of the datasheet where EMAP and FlowEHR are explained instead of keeping them in the main article
- An overall depiction of CDE and its pillars may be useful

**Response:** We're happy to take editorial guidance on this. Our inclination is to include the detailed diagrams in the main text and to leave the textual and more technical descriptions of Pillars 1 and 2 in the ESM. We think this keeps the main article readable as an overall 'sketch' but allows the interested reader to obtain further detail. To this end (as per Reviewer 3), we have added extended captions rather than extend the body text.

### 2022-07-01 rebuttals/response-reviewer-2.md

2. The figures are very nicely done and quite informative.

    Thank you again.

## Major Limitations
1. The analogy to drug discovery is not a great fit, in my opinion, and its shortcomings reveal an opportunity to better-describe the relationships between ML4H tools and the problems they seek to solve.

    **Response:** Please see the response under Major Comments (1) below.

@@

2. Some areas of the prose could benefit from increased specificity and/or clarity.

    **Response:** Please see the response under Major Comments (2) below.

@@

3. Much of the supplemental content would be better served in the main text.

    **Response:** Please see the response under Major Comments (3) below.

## Major Comments
1. The analogy to drug discovery is a poor fit for this essay. One key premise underlying drug discovery activities is that potentially-actionable targets have already been identified through prior research. For example, the decision to seek JAK2-interacting molecules emerged from the understanding that JAK2 regulates important cytokines in a number of disease states. The algorithmic equivalent of this kind of prior knowledge is not apparent to me. Without a clear understanding of (a) a well-articulated problem to be solved and (b) a reasonable mechanism/logic model in which an ML tool - provided it performs well - might affect stakeholder behavior, translational failure seems essentially guaranteed. Said differently, the bake-off between different ML approaches can really only be held once the model ("ligand")'s "receptor" has been mapped/described, which I would argue happens infrequently at present. As I read the authors' proposed pillars, I found myself reflecting on the need for this kind of formative work, informed by stakeholders, to ensure that the CDE can be used for properly-specified development, operations, and evaluation.
I might suggest omitting or significantly abbreviating this section, which could open space to insert valuable information from the supplement. I also think a paragraph commenting on the relationship of the CDE to the organization's "problem identification and specification" apparatus would be a very helpful "bookend" to the implementation and continuous evaluation portions of the essay.
**Response:**
We thank the reviewer for this comment but feel that changing the manuscript as they suggest would not be the right thing to do. The point of the drug development pipeline analogy is to articulate the complexity of the intellectual work required for innovation. No-one expects a drug to be developed from scratch in a 3-5 years. However, in many cases people are disappointed when healthcare AI algorithms do not yield practice-changing results in the timeframe of a single project. The benefit of the Drug Trial Phases is that people are easily able to understand the risk level, the size of trial and the appropriate endpoints. This same thinking needs to be related to algorithms. Similar points have been made by Pepe (2005) and Ferrante di Ruffano with respect to biomarker discovery.
To specifically address the JAK-2 example, this work would happen in the pre-clinical phase that we describe as algorithm discovery. The AI equivalent of JAK-2 (a targetable molecule) is a feature (a representation of data that has some utility in classification or prediction that also has a causal relationship with the phenomenon under study)

2. Some areas of the prose could benefit from increased specificity and/or clarity. One of the challenges in the healthcare data science space is striking the right balance of technical language to meet a readership with different levels of familiarity to clinical medicine, coding/development, and operations. Some areas of the prose - e.g., lines 63-67 - skew a bit heavy into the data science jargon and might benefit from additional explanations or a bit less jargon. It may be worth the authors having a colleague with less technical familiarity provide some informal copy-editing to help with readability for an audience of varying levels of expertise.
**Response:**
Thank you. This was helpful. I have asked two colleagues to separately proof the manuscript. They have made a series of small stylistic changes that I have incorporated into the manuscript.
e.g. lines 63-67
Before:
> Given the sensitive nature of health data, these initiatives depend on expanding investment into Trusted Research Environments (TRE). [@2021c] Data flows from source (primary, secondary, social care and elsewhere) to a single secure landing zone where research teams write the code to link, clean and analyse the data. The insights return to the bedside through clinical guidelines and policy. This offline '_data-to-code_' approach is also the dominant design pattern in ML4H projects but is fundamentally flawed.
After:
> Given the sensitive nature of health data, these programmes depend on Data Safe Havens (also known as Trusted Research Environments [TRE]). [@2021c] TREs are an example of 'data-to-modeler' (DTM) designs where source data (from primary, secondary, and social care) **moves to** a separate, secure landing zone. Here research teams link, clean and analyse the data. Their insights eventually return to the bedside through clinical guidelines and policy. In ML4H, DTM is also the dominant design pattern but is fundamentally flawed.

    NB:
    A detailed diff of these changes is available [here](https://github.com/docsteveharris/2022-adversarial-penguin/pull/7)

3. The supplement - especially items 1 and 2 - are very nice and highly practical. Some elements of these items could serve to address issue 2 above. Further, the ML-Ops design principles are particularly helpful explicating how code-to-data can address the limitations of data-to-code. I think it would be great if these could be a bit more prominent, ideally in the main manuscript.
**Response:**
Thank you. Our idea was to keep the main text narrative and then allow the interested reader to use the ESM for details with the figures providing the bridge. To this end we have (partly in response to Reviewe 3) expanded the captions so the flow of the text remains roughly the same but there is additional detail at hand.

## Minor Comments
1. The manuscript has a number of instances with unspecified pronouns, e.g., sentences starting with "This" or "That," which do not specify what "This" or "That" actually are.  
**Response:** Thank you. We hope we have corrected this with the proof reading and recommendations (esp. reviewer 1).

2. Supplement page 2, typo in the ML-0PS header (zero instead of O)  
**Response:** Thanks for this careful proofread. I double checked. I think this is a quirk of the font that Overleaf is using for the online submission and I can confirm the source document uses `O` not `0`.
i.e.
```markdown
# ML-Ops (for health)
```
not
```markdown
# ML-0ps (for health)
```
